What is an Ancestral Recombination Graph?

### 2022-02-13 paper.tex
First few paragraphs of "our model" section

The deep connection between the ARG and the pedigree is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions of the ARG have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of the ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the expressivity of the structure.
Following \citep{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
\caption{\label{fig-pedigree-and-arg}
The (unannotated) ARG is closely related to the pedigree
graph, where each ``individual'' node is replaced by two ``genome''
nodes, and edges between parents and children are only included if there is
transfer of ancestral material. We usually only include nodes in an ARG
that affect our sample, so that, for example, individuals that are ancestral
to only one sampled genome are not included.}
\end{figure}

Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume a diploidy here for concreteness, but there
is no inherent limitation on ploidy in the model).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inheritied directly from $I_4$'s genome $G$ without
recombination.

[Here are the basic components of the model]

\begin{enumerate}
\item A node corresponds to one of a diploid individual's two genomes.
This is linked explicitly to the pedigree, and is fully general.
Nodes \emph{may} be labelled with certain flags describing their properties,
but it's not necessary.
\item An edge between two nodes defines a genetic inheritance relationship
between two nodes. Edges are annotated with information about the
passage of ancestral materal. There are different forms these annotations
can take. For example, we can follow the approach taken in the Griffiths
ARG and just annotate recombinant edges with the interval $[0, x)$ or $[x, L)$,
depending on side of the breakpoint the edge is on. In the most
general case, we annotate each edge with the full set of disjoint
intervals of ancestral material that flows along that edge.
\citep{mcgill2013graphml} also proposed full ancestral material being
associated with each edge.
\end{enumerate}

\begin{figure}
\vspace{5em}
% NB: not clear if this figure is better off on its own or should be
% merged with the pedigree one.
\caption{\label{fig-arg-annotations}
The specific path through the ARG for a given genome position (and therefore
the local genealogical tree) cannot be determined without annotations
to the graph. (A) The classical Griffiths approach annotates recombination
nodes with the corresponding breakpoint. (B) We can equivalently
annotate the \emph{edges} with the genomic intervals carrying ancestral
material. Edge annotations are somewhat more general, and have
significant computational advantages.}
\end{figure}

@@

\section*{Relationship with other types of phylogenetic networks}

>>>>

The deep connection between the ARG and the pedigree is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions of the ARG have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of the ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the expressivity of the structure.
Following \cite{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
    \includegraphics[width=\textwidth]{illustrations/pedigree_figure}
\caption{\label{fig-pedigree-and-arg}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined as arbitrary inheritance
of genomes between diploid (or any ploidy; see text) individuals.
(A) Each individual consists of two genomes, and each genome is
inherited as a recombinant mixture of the individual's parent's
genomes. (B) Genomes are modelled as nodes in a graph, and the edges
denote inheritance. Each edge is annotated with genome coordinates
defining the specific region(s) inherited. An ARG is a retrospecitive
structure defined with
respect to some set of sample genomes (A--D here), and so only
genomes directly ancestral to those samples are included.
(C) This description of inheritance fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

Fig~\ref{fig-pedigree-and-arg} illustrates the basic elements of the model.
The pedigree is composed of a set of individuals who each carry
two \emph{genomes} (we assume diploidy here for concreteness, but there
is no inherent limitation on ploidy in the model).
Each genome is inherited from one of the individual's parents, and may
be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes $A$ and $B$,
inherited from parents $I_3$ and $I_4$. Genome $A$ is the product of
recombining $I_3$'s two genomes $E$ and $F$ at position FIXME,
whereas $B$ was inherited directly from $I_4$'s genome $G$ without
recombination.

@@

\textcolor{red}{TODO: reduce this down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature.}
